---
title: "Calculating Bobcat Life Table, Survival and Growth"
author: "Nikole Vannest, Teague Tran, Alyssa Kibbe"
date: "2/8/2022"
output: html_document
---

## The goal of this Rmd is to:

1.  Calculate the survivorship of Bobcat age classes from population data.
2.  Model survivorship to represent stochasticity by sampling survival values between each age class.
3.  Determine survivorship under hunting, road pressure and low pressure (using both hunting and road pressure) to represent environmental pressures.
4.  Run all four scenarios (no pressure, hunting, roads and both) with stochastic values (100 model iterations) as life tables to determine the growth rate for each iteration.
5.  Visualize survivorship and variability in growth rate.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
library(here)
library(kableExtra)
library(cowplot)
library(dataRetrieval)

options(scipen=999)  # Disable Scientific Notation
```

### Start adding life table data here

```{r read in data}
bobcat_data <- read.csv(here("data", "bobcats_lifetable.csv"), header = T)
```

### Calculate Survivorship

```{r convert to survivorship}
# create empty list
surv_data <- c()

# fill list with survival proportions via loop
for(i in 1:nrow(bobcat_data)){
  surv_data[i] <- bobcat_data$Counts[i]/bobcat_data$Counts[1]
}

# Add to bobcat_data data frame
bobcat_data$survival <- surv_data
colnames(bobcat_data) <- c("age", 
                           "mx",
                           "count",
                           "survival")
```

### Fit survival to a curve

```{r fit survival to curve}
# smooth survival data as curve

ggplot(data = bobcat_data,
       aes(x = age,
       y = survival))+
  geom_point()

plot(bobcat_data$age, bobcat_data$survival,
     main="Bobcat Survival Analysis",
     xlab = "Age (years)",
     ylab = "Survival (proportion)")

#making the best fit line from our data
line_fit = lm(log(bobcat_data$survival)~ #only take the log(y)
                bobcat_data$age)
summary(line_fit)

factor = line_fit$coefficients["bobcat_data$age"]
coeff = exp(line_fit$coefficients["(Intercept)"])

fit_eq = paste("y=",coeff, "exp(",factor,"x)")
#text(20,10,fit_eq)
#adds to the plot above
#create function called survivorship
survivorship <- function(x){
  coeff*exp(factor*x)
}
#add to plot above
curve(survivorship,0,14, add = TRUE, col="blue")
```

### Create list of survival data points to sample from

```{r}
# loop through survival to make list of potential points

# Remove the first row
bobcat_survival <- bobcat_data[-1,]

# Create list to loop into
list_survival <- list()

# Use loop to create list of lists containing potential survival data from age a to b
for(i in 1:13){
a <- seq(from = bobcat_data$survival[i], to = bobcat_data$survival[i+1], by = -0.001)
list_survival[[i]] <- a
}

# Create multiple life tables with sampled survival data
survival_sampled <- list()
for(a in 1:13){
  for(i in length(list_survival[[a]])){
    survival_sampled[[a]] <- sample(list_survival[[a]], size = 100, replace = T)
  }
}

#create data frame for sampled survival data

survival_sampled_df <- data.frame(survival_sampled) 

#give it column names
colnames(survival_sampled_df) <- c("0_1",
                            "1_2", 
                           "2_3",
                           "3_4",
                           "4_5",
                           "5_6",
                           "6_7",
                           "7_8",
                           "8_9",
                           "9_10",
                           "10_11",
                           "11_12",
                           "12_13"
                           )
# transpose table so that it'll look like the bobcats dataframe
transposed_df<- t(survival_sampled_df)

#combine data column 
bobcat_lifetable <- cbind(bobcat_survival, transposed_df)
```

### Reinsert 0-year survival

```{r Insert first row of 1s}
# create sequence to reinsert 1s for year 0 in first row
ones <- seq(to = 1, from = 1, length.out = ncol(bobcat_lifetable))

# add correct data to the first 3 columns
ones[1:3] <- c(0, 0, 1517)

# bind to the bobcat_lifetable
bobcat_lifetable <- rbind(ones, bobcat_lifetable)
```

### Calculate hunting and roadkill survival

```{r calculate hunting and roadkill survival}
# Use bobcat survival and subtract hunting and road kill mortality pressure

hunting_low <- 0.1 #low hunting pressure
hunting_high <- 0.3 #high hunting pressure

roadkill_low <- 0.06 #low roadkill pressure
roadkill_high <- 0.2 #high roadkill pressure

#subtract low pressures from bobcat survival data
low_pressure_survival <- bobcat_lifetable[,5:104]-(hunting_low+roadkill_low)
low_pressure_survival <- cbind(bobcat_lifetable[,1:4], low_pressure_survival)

#replace first row with 1s
low_pressure_survival[1, 5:104] <- 1

#remove negative data points
low_pressure_survival[low_pressure_survival<0] <- 0

# #subract high pressures from bobact survival data
# high_pressure_survival <- bobcat_lifetable[,5:104]-(hunting_high+roadkill_high)
# high_pressure_survival <- cbind(bobcat_lifetable[,1:4], high_pressure_survival)
# 
# #remove negative data points
# high_pressure_survival[high_pressure_survival<0] <- 0

#subract only road kill pressure
road_pressure <- bobcat_lifetable[,5:104]-(roadkill_low)
road_pressure <- cbind(bobcat_lifetable[,1:4], road_pressure)

#replace first row with 1s
road_pressure[1, 5:104] <- 1

#remove negative data points
road_pressure[road_pressure<0] <- 0

#subract only road kill pressure
hunting_pressure <- bobcat_lifetable[,5:104]-(hunting_low)
hunting_pressure <- cbind(bobcat_lifetable[,1:4], hunting_pressure)

#replace first row with 1s
hunting_pressure[1, 5:104] <- 1

#remove negative data points
hunting_pressure[hunting_pressure<0] <- 0
```

### Plotting the different survival rates

```{r Visualize average survival}
mean_survival <- data.frame('age' = bobcat_lifetable$age,
                            "noPressure" = rowMeans(bobcat_lifetable[5:104]),
                            "lowPressure" = rowMeans(low_pressure_survival[5:104]),
                            "lowHunting" = rowMeans(hunting_pressure[5:104]),
                            "lowRoads" = rowMeans(road_pressure[5:104]))

mean_longer <- mean_survival %>% pivot_longer(cols = c('noPressure', 'lowPressure', 'lowHunting', 'lowRoads'),
                                              names_to = 'pressureType')

plot_avg_survival <- mean_longer %>% ggplot(aes(x = age, color = pressureType)) + 
  geom_point(aes(y = value, 
                 shape = pressureType),
             size = 3) +
  xlab("Age (years)") +
  ylab("Survival") +
  ggtitle("Average Survival Rates with Stochastic Variability") +
  theme(text = element_text(size = 16))

plot_avg_survival

ggsave("CompareAverageSurvival.png", plot = plot_avg_survival)
```

# Creating Lists of Average Growth Values

### Create Life Table with No Pressure

```{r Create Life Table}
#create iteration to calculate r; we are only interested in the R values! 
rlist <- list()
for(i in 5:ncol(bobcat_lifetable)) { #code to iterate across columns; goes from fifth column until it finishes!
  lxmx <- bobcat_lifetable$mx* bobcat_lifetable[ , i] 
  R0 <- sum(lxmx)
  G <- sum(bobcat_lifetable$age*lxmx)/R0
  approx.r <- log(R0)/G
  r <- mean(approx.r) #approximated growth rate from df
  y <- sum(exp(-r*bobcat_lifetable$age)*lxmx) # 
  
while (abs(y-1) >= 0.000001) {
  if (y-1>0){
    r <- r+0.00000001
  }
  else{
    r <- r-0.00000001
  }
  y <- sum(exp(-r*bobcat_lifetable$age)*lxmx)
  r
}
  rlist <- append(rlist, r) #create a list of r values from the stuff; a range of these values will give you the good good
  
}
```

### Incorporate Low hunting and roadkill pressure to life table

```{r incorporate hunting and road kill to get growth rate for low_pressure}
#create iteration to calculate r; we are only interested in the R values! 
rlist_low_pressure <- list()
for(i in 5:ncol(low_pressure_survival)) { #code to iterate across columns; goes from fifth column until it finishes!
  lxmx <- low_pressure_survival$mx* low_pressure_survival[ , i] 
  R0 <- sum(lxmx)
  G <- sum(low_pressure_survival$age*lxmx)/R0
  approx.r <- log(R0)/G
  r <- mean(approx.r) #approximated growth rate from df
  y <- sum(exp(-r*low_pressure_survival$age)*lxmx) # 
  
while (abs(y-1) >= 0.000001) {
  if (y-1>0){
    r <- r+0.00000001
  }
  else{
    r <- r-0.00000001
  }
  y <- sum(exp(-r*low_pressure_survival$age)*lxmx)
  r
}
  rlist_low_pressure <- append(rlist_low_pressure, r) #create a list of r values from the stuff; a range of these values will give you the good good
  
}

```

### Just hunting pressure

```{r hunting pressure only}
#create iteration to calculate r; we are only interested in the R values! 
rlist_hunting <- list()
for(i in 5:ncol(hunting_pressure)) { #code to iterate across columns; goes from fifth column until it finishes!
  lxmx <- hunting_pressure$mx* hunting_pressure[ , i] 
  R0 <- sum(lxmx)
  G <- sum(hunting_pressure$age*lxmx)/R0
  approx.r <- log(R0)/G
  r <- mean(approx.r) #approximated growth rate from df
  y <- sum(exp(-r*hunting_pressure$age)*lxmx) # 
  
while (abs(y-1) >= 0.000001) {
  if (y-1>0){
    r <- r+0.00000001
  }
  else{
    r <- r-0.00000001
  }
  y <- sum(exp(-r*hunting_pressure$age)*lxmx)
  r
}
  rlist_hunting <- append(rlist_hunting, r) #create a list of r values from the stuff; a range of these values will give you the good good
  
}
```

### Just road kill pressure

```{r roadkill pressure only}
#create iteration to calculate r; we are only interested in the R values! 
rlist_roadkill <- list()
for(i in 5:ncol(road_pressure)) { #code to iterate across columns; goes from fifth column until it finishes!
  lxmx <- road_pressure$mx* road_pressure[ , i] 
  R0 <- sum(lxmx)
  G <- sum(road_pressure$age*lxmx)/R0
  approx.r <- log(R0)/G
  r <- mean(approx.r) #approximated growth rate from df
  y <- sum(exp(-r*road_pressure$age)*lxmx) # 
  
while (abs(y-1) >= 0.000001) {
  if (y-1>0){
    r <- r+0.00000001
  }
  else{
    r <- r-0.00000001
  }
  y <- sum(exp(-r*road_pressure$age)*lxmx)
  r
}
  rlist_roadkill <- append(rlist_roadkill, r) #create a list of r values from the stuff; a range of these values will give you the good good
  
}

```

### Create graph to display growth rates

```{r Growth rate visual}
#Combine rlists into data frame 
r_df <- data.frame("modelRun" = c(1:100),
                   "noEffect" = as.numeric(rlist),
           "lowPressure" = as.numeric(rlist_low_pressure),
           "huntingPressure" = as.numeric(rlist_hunting),
           "roadPressure" = as.numeric(rlist_roadkill))

names(r_df) <- c("modelRuns", "None","Low (16%)", "Low Hunting (10%)", "Low Road (6%)")

# Convert it to long format so can use facet_wrap
pressure_df <- r_df %>% pivot_longer(cols = c("None",
                                              "Low (16%)",
                                              "Low Hunting (10%)", 
                                              "Low Road (6%)"), 
                                     names_to = "pressureType")

pressure_plot <- pressure_df %>% ggplot(aes(y = value, fill = pressureType)) + 
  stat_boxplot(geom = 'errorbar', width = 0.15) +
  geom_boxplot() +
  facet_wrap(~pressureType, nrow = 1, ncol = 4) +
  xlab("Pressure Type") +
  ylab("Growth Rate") +
  theme(legend.position = "none",
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        text = element_text(size = 12))

pressure_plot

ggsave("pressure_output.png", plot = pressure_plot)

```
